 

### 2021-01-19 vertical-thermal-review.Rmd

Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
There is also increasing severity and frequency of heat waves that will potentially reduce [forest] biomass acculumation, and disrupt photosynthesis and metabolic processes across plant species [Sullivan et al. 2020, @perkinsIncreasingFrequencyIntensity2012].
These changes are expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].

@@

The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019], each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019]. In wetter conditions, available water facilitates leaf transpiration through open stomata; however increased relative humdity in tropical regions indirectly caused by increased temperature reduces transpiration, leading to high leaf temperatures in the canopy [Tibbitts 1979; Perez and Feeley 2018].

Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020]. Larger trees suffer hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al. 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a]. Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020]. However, in mesic forest conditions, availability of soil water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration even during increasing air temperature, thereby buffering understory microclimates, unless relative humidity is so high as to reduce rate of evapotranspiration [Perez and Feeley, 2018].

While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020]. Understories experience a distinct set of pressures as microclimatic warming combines with conditions of limited light and related foliar traits, inter-species resource competition, and herbivory [@niinemetsResponsesForestTrees2010a; @bartholomewSmallTropicalForest2020]. Since temperature sensitivity is directly linked with plant metabolic processes and performance, even around 1 degree of warming can lead to changes in plant species composition and growth [@bertrandChangesPlantCommunity2011; @bertrandCommentForestMicroclimate2020]. Plant communities in forest microclimates are already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013]. Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020]. In addition, understory growth phenology is sensitive to warming year round, and especially in spring [Zohnner_Renner 2019].

@@

Tropical forests with densely packed overstory trees limit light to subsequent canopy layers and forest understories. Tropical forest floors receive approximately 1-2%  of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006]. However, vertical canopy layers and forest floors also receive light permeating through canopy gaps, as well as intermittent light penetration from tree, branch, and leaf movement caused by wind, defined collectively as sunflecks. Sunflecks can provide 10-80% of photon flux density for photosynthesis in canopy layers and understory, depending on canopy attributes such as foliage density, height and architecture [Way and Pearcy 2012, REF]

This difference in light regimes along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity. Leaves thus exhibit variation in leaf trait adaptations from sun to shade both along a vertical intra-canopy gradient, as well as inter-canopy [@michaletzEnergeticCarbonEconomic2016; @fauset_differences_2018; @niinemetsResponsesForestTrees2010, @sackHowStrongIntracanopy2006; @doughtyAreTropicalForests2008; @niinemetsControversyTraitsConferring2006].

@@

Wind speed in the upper canopy of a semi-deciduous tropical forest has been observed to be higher in the dry season than the wet season [@rey-sanchez_spatial_2016] [*double check*]. 
Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a]. Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{air}$ and $T_{leaf}$, which is more typical of small leaves in the upper canopy, allowing them to dissipate heat more efficiteny than larger leaves [@bauerleSeparatingFoliarPhysiology2011]. In low wind conditions, because $g_{a}$ to heat transfer is comparitevely small, $T_{leaf}$ in the upper canopy can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; Leigh et al. 2017]. Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy? [ref]

@@

RH, in combination with temperature, determines vapor pressure deficit, $VPD$. Atlantic forests in Brazil during the wet season tend to have higher $T_{air}$, and $VPD$ in the upper canopy and canopy gaps than lower in the understory where $T_{air}$ and $VPD$ tend to be lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].  
This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomatal constrains and lower osmotic potential. [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]. In northeastern US temperate forests, the upper canopy was observed to be more sensitive to $VPD$, whereas the understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]

**Finally, carbon dioxide (CO~2~) concentrations tend to be higher in the understory.** 
	Higher in the understory, particularly at dusk [@koike_leaf_2001]. Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999]. Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]

@@

If soil and stem respiration do not acclimate fully to rising temperatures, the greater amount of CO~2~ emitted at the forest floor (i.e. from soils and decaying organic matter and leaf litter) could increase the vertical CO~2~ gradient in the future, but this would still only be a short-term effect in the early morning, as the CO~2~ would quickly diffuse through the canopy. Furthermore, increases in mortality of big trees (e.g. Senf et al. 2018) would increase canopy roughness and prevent build-up of CO~2~ concentration gradients.
(Elsa: see these papers on the long-term Harvard Forest soil warming experiments. CO2 emissions increased, then decreased, then increased again. It's thought that the changes in soil CO2 emissions over time were linked to turnover in microbial community composition.
- DeAngelis, K.M., Pold, G., Topçuoğlu, B.D., van Diepen, L.T., Varney, R.M., Blanchard, J.L., Melillo, J. and Frey, S.D., 2015. Long-term forest soil warming alters microbial communities in temperate forest soils. Frontiers in microbiology, 6, p.104.
- Melillo, J.M., Steudler, P.A., Aber, J.D., Newkirk, K., Lux, H., Bowles, F.P., Catricala, C., Magill, A., Ahrens, T. and Morrisseau, S., 2002. Soil warming and carbon-cycle feedbacks to the climate system. Science, 298(5601), pp.2173-2176.
- Melillo, J.M., Frey, S.D., DeAngelis, K.M., Werner, W.J., Bernard, M.J., Bowles, F.P., Pold, G., Knorr, M.A. and Grandy, A.S., 2017. Long-term pattern and magnitude of soil carbon feedback to the climate system in a warming world. Science, 358(6359), pp.101-105.) 
- Nottingham, A.T., Meir, P., Velasquez, E. and Turner, B.L., 2020. Soil carbon loss by experimental warming in a tropical forest. Nature, 584(7820), pp.234-237.

@@

Species that have higher shade casting ability such as braodleaved trees increase T buffering [@zellweger_seasonal_2019,@vonarxSpatiotemporalEffectsForest2012]
The strength of buffering also varies with respect to geographic and climatic factors.
* Distance to coast, topographic position, elevation [@zellweger_seasonal_2019]
* [@davis_microclimatic_2019]--Proximity to riparian areas tends to sustain buffering, protecting understory temperatures even during increasing temperatures [@davis_microclimatic_2019]
However vertical climate gradients within forests are known to be steeper than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017; @martinCovarianceSunShade2020a]

@@

Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [Rosendaal et al. 2006]. After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [Keenan and Niinemets et al.2016; @niinemetsLeafAgeDependent2016a]. A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR, and the ratio of red to far red radiation [R:FR] intercepting the leaf surface. This enables acclimation or avoidance responses through the development of necessary leaf traits [Casal, 2013]. Degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], ultimately influencing leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al. 2016].

@@

Taller trees in dense forests tend to be exposed to greater irradiance but maintain shaded leaves in canopy interior and lower canopy, while shorter understory trees are shaded by overstory. As a result of different light regimes, a sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, and hairier, with thicker cuticle and greater lobation than shade leaves
[Rozendaal et al. 2006, Mathur et al. 2018, more REFS]."Additionally, shorter understory trees that spend the bulk of their lifecycle in the shade experience different pressures than plant that get the bulk of their carbon from sun-exposed leaves, but can maintaing a few layers of shade leaves as well" -Martijn

The combination of stressors experienced by taller trees' foliage are different from that experienced by understory trees. For example, overstory sun leaves experience greater loads of PAR, $VPD$, elevated temperatures, higher wind speed, etc.. whereas understory leaves are subjected to shade or limitation of light, niche competition and herbivory. Overstory sun leaves are subjected to overheating pressures; thus, through developed traits of increased stomatal density, transpiration and water supply, along with increased convective heatloss via reduced boundary layer resistance, sun leaves are able to maintain optimum temperatures for photosynthesis and metabolic processes. Additionally, overstory sun leaves tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA. Adaptive traits in understory enable light capture efficiency while maintaining larger width leaves with comparatively lower stomatal density due to available water, lesser $VPD$, and wind speed. Those differences in traits are likely to affect their thermal properties as well. Leaves act as windows to plant performance, enhancing plant adaptation to changing environment, thus leaf traits acclimated to prior stressors have been suggested to enable whole plants to respond differently to forthcoming stress [Rozendaal et al. 2006; @niinemetsResponsesForestTrees2010a]

However, even with distinct features of sun and shade traits, trait adaptation to light maximizes light use efficiency for photosynthesis at both high and low light, while also adapting to other stressors as mentioned above. In light limited conditions, shade leaves increase light capture efficiency by increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis, which "leads to a similar chlorphyll content on an area basis compared with sun leaves. This enables shade leaves to realize similar light capture to sun leaves at a lower biomass investment" [Chazdon et al. 1996, Poorter et al. 2000, Rozendaal et al. 2006]

With increasing height and age, competition for light decreases as availability  of light increases in dense forests. However, depending on growth strategies of species occupying the understory, leaf traits might differ due to species inherent light requirements. Early successional species are considered more shade intolerant than late successional species that are more shade tolerant. Ontogenetic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$), greater photosynthetic capacity in low light and greater carbon gain capacities in shade tolerators than in intolerators [Niinemets, 2006]. Shade intolerant species experience greater heterogeniety in light during ontogeny [why?-nidhi],thus LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light [H. Poorter, U. Niinemets, L. Poorter, I. J. Wright, R. Villar, Causes and consequences of variation in leaf mass per area (LMA): A meta-analysis. New Phytol. 182, 565–588 (2009).].Rozendaal et al. 2006 found no significant trait differences with ontogeny and adult stature trees. Koike et al. 2001 has more info about this.

@@

Several studies point towards common variations and differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. The differences among these traits contribute to leaf heat dissapation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table 1. summarizes relevant traits for thermal sensitivity.

@@

Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light. However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area- based respiration versus foliar nutrients, where as Q10 and E10 remained contant with height [Cavaleri et al. 2008]. In temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [Coble & Cavaleri 2015; Coble et al. 2016].In addition, carbon isotope composition also increased with height, independent of ligtht [Coble & Cavaleri 2015]

Height-based leaf trait variation in Quercus alba and Quercus velutina showed decrease in lobation with increased height in two tall species, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020]. However, Sack et al. 2006 did not observe trait variation based on height in Quercus spp.

@@

Leaf temperature is generally assumed to increase with leaf area ($LA$) [ Wright et al. 2019]. Yet, thermal imagery shows that though leaf area can be a predictor for leaf temperatures ranges [$Trange$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017]. With increase in efffective leaf width, boundary layer thickness and resistance increases along with stomtal resistance [@leighInfluenceLeafSize2017; Wright et al. 2019; @fauset_differences_2018].This lowers evaporation from stomatal pores, and in turn transpirational cooling in large leaves [@fauset_differences_2018]. Because the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissapate heat more efficiently than large leaves [@leighInfluenceLeafSize2017]. Using leaf energy balance to compare leaf temperatures of two species $Miconia$ and $Alchornea$, Fauset et al. 2016 found leaf width to be a dominant factor in determining changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$. Leaf width and characteristic dimension is generally greater in shaded leaves along the vertical gradient than exposed upper canopy [ref]. Additionally, leaf characteristics such as lobes, leaf elongation and serration, which break the entirety of leaf dimension by enhancing a smaller characteristic dimension, all facilitate more efficient heat dissipation. Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies [Vogel 1970; Roth-Nebelsick 2001;Schuepp 1993; Sack et al. 2006].However, Perez and Feeley et al. 2020 showed only a slight negative effect on leaf temperature with leaf width.

@@

incr.humidity = decreased VPD = dec. latent heat loss = limited c assimilation [Perez et al. 2020]

>>>>

Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
There is also increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012] that will potentially reduce [forest] biomass  accumulation, and disrupt photosynthesis and metabolic processes across plant species [@sullivanLongtermThermalSensitivity2020].
These changes are expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].

@@

The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019], each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019]. In wetter conditions, available water facilitates leaf transpiration through open stomata; however increased relative humidity in tropical regions indirectly caused by increased temperature reduces transpiration, leading to high leaf temperatures in the canopy [@tibbittsHumidityPlants1979; @perezIncreasingHumidityThreatens2018]

Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020]. Larger trees suffer hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; Brienen et al. 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a]. Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020]. However, in mesic forest conditions, availability of soil water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration even during increasing air temperature, thereby buffering understory microclimates, unless relative humidity is so high as to reduce rate of evapotranspiration [@perezIncreasingHumidityThreatens2018].

While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020]. Understories experience a distinct set of pressures as microclimatic warming combines with conditions of limited light and related foliar traits, inter-species resource competition, and herbivory [@niinemetsResponsesForestTrees2010a; @bartholomewSmallTropicalForest2020]. Since temperature sensitivity is directly linked with plant metabolic processes and performance, even around 1 degree of warming can lead to changes in plant species composition and growth [@bertrandChangesPlantCommunity2011; @bertrandCommentForestMicroclimate2020]. Plant communities in forest microclimates are already undergoing thermophilization [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013]. Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020]. In addition, understory growth phenology is more sensitive to warming than canopy-- showing earlier spring onset and later autumn senescense [@zohnerOngoingSeasonallyUneven2019].

@@

Tropical forests with densely packed overstory trees limit light to subsequent canopy layers and forest understories. Tropical forest floors receive approximately 1-2%  of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006]. However, vertical canopy layers and forest floors also receive light permeating through canopy gaps, as well as intermittent light penetration from tree, branch, and leaf movement caused by wind, defined collectively as sunflecks. Sunflecks can provide 10-80% of photon flux density for photosynthesis in canopy layers and understory, depending on canopy attributes such as foliage density, height and architecture [@waySunflecksTreesForests2012]

This difference in light regimes along a vertical profile is an important driver of plasticity in leaf traits, leaf physiology, adaptation, and thermal sensitivity. Leaves thus exhibit variation in leaf trait adaptations from sun to shade both along a vertical intra-canopy gradient, as well as inter-canopy [@michaletzEnergeticCarbonEconomic2016; @fauset_differences_2018; @niinemetsResponsesForestTrees2010; @sackHowStrongIntracanopy2006; @niinemetsControversyTraitsConferring2006].

@@

Wind speed in the upper canopy of a semi-deciduous tropical forest has been observed to be higher in the dry season than the wet season [@rey-sanchez_spatial_2016] [*double check*]. 
Greater upper canopy wind speeds result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a]. Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{air}$ and $T_{leaf}$, which is more typical of small leaves in the upper canopy, allowing them to dissipate heat more efficiently than larger leaves [@bauerleSeparatingFoliarPhysiology2011]. In low wind conditions, because $g_{a}$ to heat transfer is comparatively small, $T_{leaf}$ in the upper canopy can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; @leighInfluenceLeafSize2017]. Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy? [ref]

@@

RH, in combination with temperature, determines vapor pressure deficit, $VPD$. Atlantic forests in Brazil during the wet season tend to have higher $T_{air}$, and $VPD$ in the upper canopy and canopy gaps than lower in the understory where $T_{air}$ and $VPD$ tend to be lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].  
This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomatal constraints and lower osmotic potential [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]. In northeastern US temperate forests, the upper canopy was observed to be more sensitive to $VPD$, whereas the understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]

**Finally, carbon dioxide (CO~2~) concentrations tend to be higher in the understory.** 
	Higher in the understory, particularly at dusk [@koike_leaf_2001]. Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999]. Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]

@@

If soil and stem respiration do not acclimate fully to rising temperatures, the greater amount of CO~2~ emitted at the forest floor (i.e. from soils and decaying organic matter and leaf litter) could increase the vertical CO~2~ gradient in the future, but this would still only be a short-term effect in the early morning, as the CO~2~ would quickly diffuse through the canopy. Furthermore, increases in mortality of big trees (e.g. @senfCanopyMortalityHas2018a) would increase canopy roughness and prevent build-up of CO~2~ concentration gradients.
(Elsa: see these papers on the long-term Harvard Forest soil warming experiments. CO2 emissions increased, then decreased, then increased again. It's thought that the changes in soil CO2 emissions over time were linked to turnover in microbial community composition.
- DeAngelis, K.M., Pold, G., Topçuoğlu, B.D., van Diepen, L.T., Varney, R.M., Blanchard, J.L., Melillo, J. and Frey, S.D., 2015. Long-term forest soil warming alters microbial communities in temperate forest soils. Frontiers in microbiology, 6, p.104.
- Melillo, J.M., Steudler, P.A., Aber, J.D., Newkirk, K., Lux, H., Bowles, F.P., Catricala, C., Magill, A., Ahrens, T. and Morrisseau, S., 2002. Soil warming and carbon-cycle feedbacks to the climate system. Science, 298(5601), pp.2173-2176.
- Melillo, J.M., Frey, S.D., DeAngelis, K.M., Werner, W.J., Bernard, M.J., Bowles, F.P., Pold, G., Knorr, M.A. and Grandy, A.S., 2017. Long-term pattern and magnitude of soil carbon feedback to the climate system in a warming world. Science, 358(6359), pp.101-105.) 
- Nottingham, A.T., Meir, P., Velasquez, E. and Turner, B.L., 2020. Soil carbon loss by experimental warming in a tropical forest. Nature, 584(7820), pp.234-237.

@@

Species that have higher shade casting ability such as broadleaved trees increase T buffering [@zellweger_seasonal_2019,@vonarxSpatiotemporalEffectsForest2012]
The strength of buffering also varies with respect to geographic and climatic factors.
* Distance to coast, topographic position, elevation [@zellweger_seasonal_2019]
* [@davis_microclimatic_2019]--Proximity to riparian areas tends to sustain buffering, protecting understory temperatures even during increasing temperatures [@davis_microclimatic_2019]
However vertical climate gradients within forests are known to be steeper than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017; @martinCovarianceSunShade2020a]

@@

Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006]. After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR, and the ratio of red to far red radiation [R:FR] intercepting the leaf surface. This enables acclimation or avoidance responses through the development of necessary leaf traits [@casalPhotoreceptorSignalingNetworks2013]. Degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; @sackHowStrongIntracanopy2006], ultimately influencing leaf structural, chemical and physiological traits as well [@keenanGlobalLeafTrait2016].

@@

Taller trees in dense forests tend to be exposed to greater irradiance but maintain shaded leaves in canopy interior and lower canopy, while shorter understory trees are shaded by overstory. As a result of different light regimes, a sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, and hairier, with thicker cuticle and greater lobation than shade leaves
[Rozendaal et al. 2006, Mathur et al. 2018, more REFS]."Additionally, shorter understory trees that spend the bulk of their lifecycle in the shade experience different pressures than plant that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well" -Martijn

The combination of stressors experienced by taller trees' foliage are different from that experienced by understory trees. For example, overstory sun leaves experience greater loads of PAR, $VPD$, elevated temperatures, higher wind speed, etc.. whereas understory leaves are subjected to shade or limitation of light, niche competition and herbivory. Overstory sun leaves are subjected to overheating pressures; thus, through developed traits of increased stomatal density, transpiration and water supply, along with increased convective heatloss via reduced boundary layer resistance, sun leaves are able to maintain optimum temperatures for photosynthesis and metabolic processes. Additionally, overstory sun leaves tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA. Adaptive traits in understory enable light capture efficiency while maintaining larger width leaves with comparatively lower stomatal density due to available water, lesser $VPD$, and wind speed. Those differences in traits are likely to affect their thermal properties as well. Leaves act as windows to plant performance, enhancing plant adaptation to changing environment, thus leaf traits acclimated to prior stressors have been suggested to enable whole plants to respond differently to forthcoming stress [@rozendaalPlasticityLeafTraits2006; @niinemetsResponsesForestTrees2010a]

However, even with distinct features of sun and shade traits, trait adaptation to light maximizes light use efficiency for photosynthesis at both high and low light, while also adapting to other stressors as mentioned above. In light limited conditions, shade leaves increase light capture efficiency by increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis, which "leads to a similar chlorophyll content on an area basis compared with sun leaves. This enables shade leaves to realize similar light capture to sun leaves at a lower biomass investment" [Chazdon et al. 1996, Poorter et al. 2000, @rozendaalPlasticityLeafTraits2006]

With increasing height and age, competition for light decreases as availability  of light increases in dense forests. However, depending on growth strategies of species occupying the understory, leaf traits might differ due to species inherent light requirements. Early successional species are considered more shade intolerant than late successional species that are more shade tolerant. Ontogenetic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$), greater photosynthetic capacity in low light and greater carbon gain capacities in shade tolerators than in intolerators [@niinemetsControversyTraitsConferring2006]. Shade intolerant species experience greater heterogeneity in light during ontogeny [why?-nidhi],thus LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light [H. Poorter, U. Niinemets, L. Poorter, I. J. Wright, R. Villar, Causes and consequences of variation in leaf mass per area (LMA): A meta-analysis. New Phytol. 182, 565–588 (2009).].Rozendaal et al. 2006 found no significant trait differences with ontogeny and adult stature trees. Koike et al. 2001 has more info about this.

@@

Several studies point towards common variations and differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table 1. summarizes relevant traits for thermal sensitivity.

@@

Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light. However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area- based respiration versus foliar nutrients, whereas Q10 and E10 remained constant with height [Cavaleri et al. 2008]. In temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].In addition, carbon isotope composition also increased with height, independent of light [@cobleLightAcclimationOptimizes2015]

Height-based leaf trait variation in Quercus alba and Quercus velutina showed decrease in lobation with increased height in two tall species, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020]. However, @sackHowStrongIntracanopy2006 did not observe trait variation based on height in Quercus spp.

@@

Leaf temperature is generally assumed to increase with leaf area ($LA$) [ Wright et al. 2019]. Yet, thermal imagery shows that though leaf area can be a predictor for leaf temperatures ranges [$Trange$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017]. With increase in effective leaf width, boundary layer thickness and resistance increases along with stomatal resistance [@leighInfluenceLeafSize2017; Wright et al. 2019; @fauset_differences_2018].This lowers evaporation from stomatal pores, and in turn transpirational cooling in large leaves [@fauset_differences_2018]. Because the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissipate heat more efficiently than large leaves [@leighInfluenceLeafSize2017]. Using leaf energy balance to compare leaf temperatures of two species $Miconia$ and $Alchornea$, Fauset et al. 2016 found leaf width to be a dominant factor in determining changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$. Leaf width and characteristic dimension is generally greater in shaded leaves along the vertical gradient than exposed upper canopy [ref]. Additionally, leaf characteristics such as lobes, leaf elongation and serration, which break the entirety of leaf dimension by enhancing a smaller characteristic dimension, all facilitate more efficient heat dissipation. Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].However, Perez and Feeley et al. 2020 showed only a slight negative effect on leaf temperature with leaf width.

@@

incr.humidity = decreased VPD = dec. latent heat loss = limited c assimilation [@perezPSIIHeatTolerances2020]
